# Geometries

(thanks to Jannis Fröhlking)

## ex 3.1 

Give two examples of geometries in 2-D (flat) space that are not simple feature geometries, and create a plot of them.

```{r}
library(sf)
x1 <- st_linestring(rbind(c(0,0),c(2,2),c(0,2),c(2,0)))
x2 <- st_polygon(list(rbind(c(3,0),c(5,2),c(3,2),c(5,0),c(3,0))))
plot(c(x1,x2), col = 2:3)
st_is_simple(x1)
st_is_simple(x2)
```

## ex 3.2 

Recompute the coordinates 10.542, 0.01, 45321.789 using precision values 1, 1e3, 1e6, and 1e-2.
```{r 2) precicsion}
for(i in c(1,1e3,1e6,1e-2)) 
  print(round(i * c(10.542, 0.01, 45321.789))/i)
```

## ex 3.3 

Describe a practical problem for which an n-ary intersection would be needed.

* for a long-term set of polygons with fire extents, find the polygons that underwent
  0, 1, 2, 3, ... fires
* for a set of extents of n individual plant species, find polygons with 0, 1, ..., n species, or
  find the polygon(s) that contain a particular subset of plant species.

## ex 3.4 

How can you create a Voronoi diagram (figure 3.3) that has closed polygons for every point?

Voronoi diagrams have "open polygons", areas that extend into
infinity, for boundary points. These cannot be represented by
simple feature geometries. `st_voronoi` chooses a default (square)
polygon to limit the extent, which can be enlarged. Alternatively,
the extent can be limited using `st_intersection` on its result:

```{r}
library(sf)
par(mfrow = c(2,2))
set.seed(133331)
mp = st_multipoint(matrix(runif(20), 10))
plot(st_voronoi(mp), col = NA, border = 'black')
plot(mp, add = TRUE)
title("default extent")
e2 = st_polygon(list(rbind(c(-5,-5), c(5, -5), c(5,5), c(-5, 5), c(-5,-5))))
plot(st_voronoi(mp, envelope = e2), col = NA, border = 'black')
plot(mp, add = TRUE)
title("enlarged envelope")
e3 = st_polygon(list(rbind(c(0,0), c(1, 0), c(1, 1), c(0, 1), c(0, 0))))
v = st_voronoi(mp) %>% st_collection_extract() # pulls POLYGONs out of GC
plot(st_intersection(v, e3), col = NA, border = 'black', axes=TRUE)
plot(mp, add = TRUE)
title("smaller, intersected envelope")
```

## ex 3.5 

Give the unary measure dimension for geometries POINT Z (0 1 1), LINESTRING Z (0 0 1,1 1 2), and POLYGON Z ((0 0 0,1 0 0,1 1 0,0 0 0))

```{r unary measure dimension}
st_dimension(st_point(c(0,1,1)))
st_dimension(st_linestring(rbind(c(0,1,1),c(1,1,2))))
st_dimension(st_polygon(list(rbind(c(0,0,0),c(1,0,0),c(1,1,0),c(0,0,0)))))
```

(these are all zero-dimensional geometries because they are _points_, irrespective the number of dimensions they're defined in)

## ex 3.6 

Give the DE-9IM relation between LINESTRING(0 0,1 0) and LINESTRING(0.5 0,0.5 1); explain the individual characters.
```{r DE-9IM relation}
line_1 = st_linestring(rbind(c(0,0),c(1,0)))
line_2 = st_linestring(rbind(c(.5,0),c(.5,1)))
plot(line_1,col = "green")
plot(line_2,col = "red", add = TRUE)
st_relate(line_1, line_2)
```
The DE-9IM relation is F01FF0102

- F   Intersection of green lines interior and red lines interior is empty
- 0   Intersection of green lines interior and red lines boundary results in one point in the middle of the green line
- 1   Intersection of green lines interior and red lines exterior results in a line covering most parts of the green line
- F   Intersection of green lines boundary and red lines interior is empty
- F   Intersection of green lines boundary and red lines boundary is empty
- 0   Intersection of green lines boundary and red lines exterior results in the two boundary points of the green line
- 1   Intersection of green lines exterior and red lines interior results in a line covering most parts of the red line
- 0   Intersection of green lines exterior and red lines boundary results in the upper boundary point of the red line
- 2   Intersection of green lines exterior and red lines results in a polygonal geometry covering everything except the two lines

(the boundary of a LINESTRING is formed by its two end points)

## ex 3.7 

Can a set of simple feature polygons form a coverage? If so, under which constraints?
Yes, but I would say that the set may just contain one polygon, because simple features provide no way of assigning points on the boundary of two adjacent polygons to a single polygon.

## ex 3.8 

For the nc counties in the dataset that comes with R package sf, find the points touched by four counties.
```{r}
# read data
nc <- st_read(system.file("shape/nc.shp", package="sf"))
# get intersections
(nc_geom = st_geometry(nc))
nc_ints = st_intersection(nc_geom)
plot(nc_ints, main = "All intersections")
# Function to check class of intersection objects
get_points = function(x){
  if(class(x)[2]=="POINT")  return(x)
}
# get points
points = lapply(nc_ints, get_points)
points[sapply(points,is.null)] <- NULL
sf_points = st_sfc(points)
st_crs(sf_points) = st_crs(nc)
# get points with four neighbouring geometries (=states)
touch = st_touches(sf_points, nc_geom)
four_n = sapply(touch, function(y) which(length(y)==4))
names(four_n) = seq_along(four_n)
point_no = array(as.numeric(names(unlist(four_n))))
result = st_sfc(points[point_no])
plot(nc_geom, main = "Points touched by four counties")
plot(result, add = TRUE, col = "red", pch = 10, cex = 2)
```

A more compact way might be to search for points where counties touch
another county _only_ in a point, which can be found using `st_relate`
using a `pattern`:

```{r}
(pts = nc %>% st_relate(pattern = "****0****"))
nc %>% st_relate(pattern = "****0****") %>% lengths() %>% sum()
```
which is, as expected, four times the number of points shown in the plot above.

How can we find these points? See [here](https://edzer.github.io/UseR2017/#higher-level-operations-summarise-interpolate-aggregate-st_join):

```{r}
nc = st_geometry(nc)
s2 = sf_use_s2(FALSE) # use GEOM geometry
pts = st_intersection(nc, nc)
pts = pts[st_dimension(pts) == 0]
plot(st_geometry(nc))
plot(st_geometry(pts), add = TRUE, col = "red", pch = 10, cex = 2)
sf_use_s2(s2) # set back
```

## ex 3.9 

How would figure 3.6 look like if delta for the y-coordinate was positive?
Only cells that were fully crossed by the red line would be grey:

```{r}
library(stars)
ls = st_sf(a = 2, st_sfc(st_linestring(rbind(c(0.1, 0), c(1, .9)))))
grd = st_as_stars(st_bbox(ls), nx = 10, ny = 10, xlim = c(0, 1.0), ylim = c(0, 1),
   values = -1)
attr(grd, "dimensions")$y$delta = .1
attr(grd, "dimensions")$y$offset = 0 
r = st_rasterize(ls, grd, options = "ALL_TOUCHED=TRUE")
r[r == -1] = NA
plot(st_geometry(st_as_sf(grd)), border = 'orange', col = NA, 
     reset = FALSE, key.pos=NULL)
plot(r, axes = TRUE, add = TRUE, breaks = "equal") # ALL_TOUCHED=FALSE;
plot(ls, add = TRUE, col = "red", lwd = 2)
```

The reason is that in this case, _lower_ left corners of grid cells are part
of the cell, rather than _upper_ left corners.

